---
title: "Split Column util"
author: "Andrew Leimdorfer"
date: "4 September 2015"
output: html_document
---

This R utility splits a column into two on a given separator. 

##User Story

**As a journalist** with a dataset that has a column of data with two values combined in a string,  
**I want** to separate those two values into separate columns, 
**so that** I can work with tidy data.

Here is one example:


```{r echo=FALSE}

a = c("Jack is dad", "Jill is mum", "Hansel is son", "Gretel is daughter")
b = c("Apples", "Pears", "Chocolate", "Sweets") 

df <- data.frame(a,b)
names(df)<- c("People", "Favourite Food")

print(df)

```


The following utility assumes that: 

 * you have one column that you want to change like the "People" column above
 * you have a part of the string that you can identify as a "splitter"
 * you want a new dataset created with a different name with the data separated in two columns
 * You have the "stringr" package installed in R (*install.packages("stringr")*)
 
 You may also have decided on a new name for the two new columns (in this case "name" and "relation" might be an option). Also, you might have a name for the new dataset, although the utility will come up with some names for you as a default.
 
## Using the utility
 
1. Make sure the R file "split\_column\_on\_str.R" is in the same directory as your data.csv file
2. Load the split\_column\_on\_str.R 

```{r}
source("split_column_on_str.R")
```

Then call the "split_col()" function passing in the following arguments:

*[YOUR DATA FILE],[THE COLUMN YOU WANT TO SPLIT], [THE SEPARATOR]

[THE NEW DATA FILENAME], [NEW COLUMN NAME 1], [NEW COLUMN NAME 2]*

(The last 3 are optional. Mistakes in the first 3 will cause errors)


```{r message=FALSE}
split_col("family.csv","People", "is","family_new.csv", "name","relation")
```

That's it! You should now have a file in the same directory with the data tidies as above.


***

## How it works

First we load the stringr library

```{r message=FALSE, eval=FALSE}

library(stringr)

```

Then we create the split_col() function and set defaults for the filename and new columns if they weren't supplied.

```{r eval=FALSE}

split_col <- function(yourdata, col_to_split, splittr, filename, newcol1, newcol2){
      
      #First we do some checks and set defaults
  
      if(missing(filename)) {
            filename <- paste0("new_", yourdata)
      }
      if(missing(newcol1)) {
            newcol1 <- paste0(col_to_split, "1")
      }
      if(missing(newcol2)) {
            newcol2 <- paste0(col_to_split, "2")
      }


```


Next we check that the data file exists. If it does we read it in to "data_in".

```{r eval=FALSE}
      if (!file.exists(yourdata)) {
            
            print("ERROR: Data file not found. Remember to include the .csv")
            break
      } else {
            
            data_in<-read.csv(yourdata,header=TRUE)
      }
```

Then we look for the index of the column name provided to the function that we are going to splt on. 

If the index is 0 it means the argument identifying the column is wrong.

```{r eval=FALSE}
      colidx <- which(colnames(data_in)==col_to_split) 
      
      if(length(colidx)<1){
            print("ERROR: Column not found. Remember they are case-sensitive.")
            break
      }

```


Now we create a new dataframe of just the column we want to split (d) and pass this along with the str we are using as the "splitter" to the "do\_the\_splits()" function.

```{r eval=FALSE}
      d <- as.data.frame(data_in[,colidx])
      s <- splittr
      
      tdf <- apply(d, 1, do_the_splits, s)
```

"do\_the\_splits()" returns a matrix of dataframes as we get a df for each row (ahaaaaaa!!!), so we need to cast result back to a dataframe but first we transpose the matrix.

```{r eval=FALSE}
      result <- as.data.frame(t(tdf))
```


Finally, we give the new dataframe's columns their names, remove the old unwanted column and cbind the new separated columns to the data from the original dataset.

```{r eval=FALSE}      
      names(result) <- c(newcol1,newcol2)
      
      data_in[,colidx]<-NULL
      
      result <- cbind(result,data_in) 

      print(result)
      
      write.csv(result, filename)
}
```

"do\_the\_splits()" which does the work on each row of data\_in checks for the splitter in the string and, if it exists, uses str\_split\_fixed (from stringr) to create 2 columns:

```{r eval=FALSE}
do_the_splits <- function(d,s) {
      
      if(!grep(s, d)){
            
            print("ERROR: Splitter not found in one or more columns. Please check the data.")
            break            
      }else{
            
            str_split_fixed(d, s, 2)
      }
}
```

That's it.

All suggestions welcome. The code lives here: 
